\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2021-02-03 analysis/paper/casebase-manuscript.Rmd
updated abstract, intro and lit review.. to also reflect 3 CS instead of 4

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given a covariate profile. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how the parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | In this paper, we present the **casebase** R package, a comprehensive and flexible toolkit for parametric survival analysis. We describe how the case-base framework can also be used in more complex settings: competing risks, time-varying exposure, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis of time-to-event data. We illustrate all these features through four different case studies.
  | *SRB and MT contributed equally to this work.
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}
  \newcommand{\proglang}{\textsf}
output: rticles::rjournal_article 
bibliography: casebase-manuscript.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

From the perspective of clinicians and their patients, the most relevant quantity is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances, and not the hazard ratio between a treatment and control group. Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can subsequently be used to estimate the cumulative incidence function (CIF). Using a parametric estimator of the hazard function leads to a smooth function of time; as a consequence, the CIF and the survival function estimates also vary smoothly over time.

@@

In this article, we present the \pkg{casebase} package [@casebase-package] implemented in \proglang{R} [@r-core], for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface. The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile. Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function. It also provides tools for variable selection and competing-risk analysis. In addition, we provide extensive visualization tools.

In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parametric hazard functions. We then give a short review of existing \proglang{R} packages that implement comparable features as \pkg{casebase}. Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions. This is followed by four case studies that illustrate the flexibility and capabilities of \pkg{casebase}. We show how the same framework can be used for competing risk analyses, penalized estimation, and for studies with time-dependent exposures. Finally, we end the article with a discussion of the results and of future directions.

@@

# Existing packages

@@

Parametric survival models are implemented in a handful of packages: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival} [-@survival-package]. The types of models they allow vary for each package. For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} also have the ability to model the effect of time using splines, which allows flexible modeling of the hazard function. Moreover, \pkg{flexsurv} has the ability to estimate both scale and shape parameters for a variety of parametric families. As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time). Therefore, our package allows the user to model the effect of time using splines. Also, by including interaction terms between covariates and time, it also allows users to fit (non-proportional) time-varying coefficient models. However, we do not explicitly model any shape parameter, unlike \pkg{flexsurv}.

>>>>

---
author:
  - name: Sahir Rai Bhatnagar*
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{sahir.bhatnagar@mail.mcgill.ca}
    url: http://sahirbhatnagar.com/
  - name: Maxime Turgeon*
    affiliation: 'University of Manitoba'
    address: >
      186 Dysart Road
      Winnipeg, MB, Canada R3T 2N2
    email: \email{max.turgeon@umanitoba.ca}
    url: https://maxturgeon.ca/
  - name: Jesse Islam 
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{jesse.islam@mail.mcgill.ca}
  - name: James A. Hanley
    affiliation: McGill University
    address: >
      1020 Pine Avenue West
      Montreal, QC, Canada H3A 1A2
    email: \email{james.hanley@mcgill.ca}
    url: http://www.medicine.mcgill.ca/epidemiology/hanley/
  - name: Olli Saarela
    affiliation: University of Toronto
    address: >
      Dalla Lana School of Public Health, 
      155 College Street, 6th floor, 
      Toronto, Ontario 
      M5T 3M7, Canada
    email: \email{olli.saarela@utoronto.ca}
    url: http://individual.utoronto.ca/osaarela/
title: "\\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates"
abstract: |
  | In epidemiological studies of time-to-event data, a quantity of interest to the clinician and the patient is the risk of an event given a covariate profile. However, methods relying on time matching or risk-set sampling (including Cox regression) eliminate the baseline hazard from the likelihood expression or the estimating function. The baseline hazard then needs to be estimated separately using a non-parametric approach. This leads to step-wise estimates of the cumulative incidence that are difficult to interpret. Using case-base sampling, Hanley & Miettinen (2009) explained how the parametric hazard functions can be estimated using logistic regression. Their approach naturally leads to estimates of the cumulative incidence that are smooth-in-time.
  | In this paper, we present the **casebase** R package, a comprehensive and flexible toolkit for parametric survival analysis. We then describe how the case-base framework can be used in more complex settings: non-linear functions of time, competing risks, and variable selection. Our package also includes an extensive array of visualization tools to complement the analysis of time-to-event data. We illustrate all these features through three different case studies.
  | *SRB and MT contributed equally to this work.
preamble: >
  \usepackage{amsmath}
  \usepackage{amssymb}
  \usepackage{longtable}
  \usepackage{graphicx}
  \usepackage{tabularx}
  \usepackage{float}
  \usepackage{booktabs}
  \usepackage{makecell}
  \usepackage{tabu}
  \DeclareUnicodeCharacter{2500}{-}
  \newcommand{\proglang}{\textsf}
output: rticles::rjournal_article 
bibliography: casebase-manuscript.bib 
fig_width: 12
fig_height: 9
fig_caption: true
editor_options: 
  chunk_output_type: console
---

@@

From the perspective of clinicians and their patients, the most relevant quantity is often the 5- or 10-year risk of experiencing a certain event given the patient's particular circumstances, and not the hazard ratio between a treatment and control group. Therefore, to make sound clinical decisions, it is important to accurately estimate the *full* hazard function, which can subsequently be used to estimate the CIF. Using a parametric estimator of the hazard function leads to a smooth function of time; as a consequence, the CIF and the survival function estimates also vary smoothly over time.

@@

In this article, we present the \pkg{casebase} package [@casebase-package] implemented in \proglang{R} [@r-core], for parametric survival analysis that combines the ideas of Hanley & Miettinen [-@hanley2009fitting] into a simple interface. The purpose of the \pkg{casebase} package is to provide practitioners with an easy-to-use software tool to compute a patient's risk (or cumulative incidence) of an event, conditional on a particular patient's covariate profile. Our package retains the flexibility of case-base sampling and the familiar interface of the \code{glm} function. It also provides tools for variable selection, competing-risk analysis, and dealing with time-dependent exposures. In addition, we provide functions for exploratory data analysis and visualizing the estimated quantities.

In what follows, we first recall some theoretical details on case-base sampling and its use for estimating parametric hazard functions. We then give a short review of existing \proglang{R} packages that implement comparable features as \pkg{casebase}. Next, we provide some details about the implementation of case-base sampling in our package, and we give a brief survey of its main functions. This is followed by three case studies that illustrate the flexibility and capabilities of \pkg{casebase}. We show how the same framework can be used for non-linear functions of time, competing risks and variable selection via penalized regression. Finally, we end the article with a discussion of the results and of future directions.

@@

# Comparison with existing packages

@@

Parametric survival models are implemented in a handful of packages, each differing in the underlying data distribution: \pkg{CFC} [-@mahani2015bayesian], \pkg{flexsurv} [-@flexsurv], \pkg{SmoothHazard} [-@smoothHazard], \pkg{rsptm2} [-@clements_liu], \pkg{mets} [-@scheike2014estimating], and \pkg{survival} [-@survival-package]. For example, \pkg{SmoothHazard} is limited to Weibull distributions [-@smoothHazard], whereas both \pkg{flexsurv} and \pkg{survival} allow users to supply any distribution of their choice. \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and \pkg{rstpm2} can model the effect of time using splines, which allows flexible modeling of the hazard function. As discussed above, \pkg{casebase} can model any parametric family whose log-hazard can be expressed as a linear combination of covariates (including time). Therefore, our package is more general in that it allows the user to model any linear or non-linear transformation of time including splines and higher order polynomials. Also, by including interaction terms between covariates and time, it also allows users to fit (non-proportional) time-varying coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly model any shape parameter.

### 2021-02-03 analysis/paper/casebase-manuscript.tex
updated abstract, intro and lit review.. to also reflect 3 CS instead of 4

\abstract{%
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician and the patient is the risk of an event given a
covariate profile. However, methods relying on time matching or risk-set
sampling (including Cox regression) eliminate the baseline hazard from
the likelihood expression or the estimating function. The baseline
hazard then needs to be estimated separately using a non-parametric
approach. This leads to step-wise estimates of the cumulative incidence
that are difficult to interpret. Using case-base sampling, Hanley \&
Miettinen (2009) explained how the parametric hazard functions can be
estimated using logistic regression. Their approach naturally leads to
estimates of the cumulative incidence that are smooth-in-time.\\
In this paper, we present the \textbf{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
describe how the case-base framework can also be used in more complex
settings: competing risks, time-varying exposure, and variable
selection. Our package also includes an extensive array of visualization
tools to complement the analysis of time-to-event data. We illustrate
all these features through four different case studies.\\
*SRB and MT contributed equally to this work.
}

@@

From the perspective of clinicians and their patients, the most relevant
quantity is often the 5- or 10-year risk of experiencing a certain event
given the patient's particular circumstances, and not the hazard ratio
between a treatment and control group. Therefore, to make sound clinical
decisions, it is important to accurately estimate the \emph{full} hazard
function, which can subsequently be used to estimate the cumulative
incidence function (CIF). Using a parametric estimator of the hazard
function leads to a smooth function of time; as a consequence, the CIF
and the survival function estimates also vary smoothly over time.

@@

In this article, we present the \pkg{casebase} package
\citep{casebase-package} implemented in \textsf{R} \citep{r-core}, for
parametric survival analysis that combines the ideas of Hanley \&
Miettinen \citeyearpar{hanley2009fitting} into a simple interface. The
purpose of the \pkg{casebase} package is to provide practitioners with
an easy-to-use software tool to compute a patient's risk (or cumulative
incidence) of an event, conditional on a particular patient's covariate
profile. Our package retains the flexibility of case-base sampling and
the familiar interface of the \code{glm} function. It also provides
tools for variable selection and competing-risk analysis. In addition,
we provide extensive visualization tools.

In what follows, we first recall some theoretical details on case-base
sampling and its use for estimating parametric hazard functions. We then
give a short review of existing \textsf{R} packages that implement
comparable features as \pkg{casebase}. Next, we provide some details
about the implementation of case-base sampling in our package, and we
give a brief survey of its main functions. This is followed by four case
studies that illustrate the flexibility and capabilities of
\pkg{casebase}. We show how the same framework can be used for competing
risk analyses, penalized estimation, and for studies with time-dependent
exposures. Finally, we end the article with a discussion of the results
and of future directions.

@@

\hypertarget{existing-packages}{%
\section{Existing packages}\label{existing-packages}}

@@

Parametric survival models are implemented in a handful of packages:
\pkg{CFC} \citeyearpar{mahani2015bayesian}, \pkg{flexsurv}
\citeyearpar{flexsurv}, \pkg{SmoothHazard} \citeyearpar{smoothHazard},
\pkg{rsptm2} \citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}
\citeyearpar{survival-package}. The types of models they allow vary for
each package. For example, \pkg{SmoothHazard} is limited to Weibull
distributions \citeyearpar{smoothHazard}, whereas both \pkg{flexsurv}
and \pkg{survival} allow users to supply any distribution of their
choice. Also, \pkg{flexsurv}, \pkg{smoothhazard}, \pkg{mets} and
\pkg{rstpm2} also have the ability to model the effect of time using
splines, which allows flexible modeling of the hazard function.
Moreover, \pkg{flexsurv} has the ability to estimate both scale and
shape parameters for a variety of parametric families. As discussed
above, \pkg{casebase} can model any parametric family whose log-hazard
can be expressed as a linear combination of covariates (including time).
Therefore, our package allows the user to model the effect of time using
splines. Also, by including interaction terms between covariates and
time, it also allows users to fit (non-proportional) time-varying
coefficient models. However, we do not explicitly model any shape
parameter, unlike \pkg{flexsurv}.

>>>>

\abstract{%
In epidemiological studies of time-to-event data, a quantity of interest
to the clinician and the patient is the risk of an event given a
covariate profile. However, methods relying on time matching or risk-set
sampling (including Cox regression) eliminate the baseline hazard from
the likelihood expression or the estimating function. The baseline
hazard then needs to be estimated separately using a non-parametric
approach. This leads to step-wise estimates of the cumulative incidence
that are difficult to interpret. Using case-base sampling, Hanley \&
Miettinen (2009) explained how the parametric hazard functions can be
estimated using logistic regression. Their approach naturally leads to
estimates of the cumulative incidence that are smooth-in-time.\\
In this paper, we present the \textbf{casebase} R package, a
comprehensive and flexible toolkit for parametric survival analysis. We
then describe how the case-base framework can be used in more complex
settings: non-linear functions of time, competing risks, and variable
selection. Our package also includes an extensive array of visualization
tools to complement the analysis of time-to-event data. We illustrate
all these features through three different case studies.\\
*SRB and MT contributed equally to this work.
}

@@

From the perspective of clinicians and their patients, the most relevant
quantity is often the 5- or 10-year risk of experiencing a certain event
given the patient's particular circumstances, and not the hazard ratio
between a treatment and control group. Therefore, to make sound clinical
decisions, it is important to accurately estimate the \emph{full} hazard
function, which can subsequently be used to estimate the CIF. Using a
parametric estimator of the hazard function leads to a smooth function
of time; as a consequence, the CIF and the survival function estimates
also vary smoothly over time.

@@

In this article, we present the \pkg{casebase} package
\citep{casebase-package} implemented in \textsf{R} \citep{r-core}, for
parametric survival analysis that combines the ideas of Hanley \&
Miettinen \citeyearpar{hanley2009fitting} into a simple interface. The
purpose of the \pkg{casebase} package is to provide practitioners with
an easy-to-use software tool to compute a patient's risk (or cumulative
incidence) of an event, conditional on a particular patient's covariate
profile. Our package retains the flexibility of case-base sampling and
the familiar interface of the \code{glm} function. It also provides
tools for variable selection, competing-risk analysis, and dealing with
time-dependent exposures. In addition, we provide functions for
exploratory data analysis and visualizing the estimated quantities.

In what follows, we first recall some theoretical details on case-base
sampling and its use for estimating parametric hazard functions. We then
give a short review of existing \textsf{R} packages that implement
comparable features as \pkg{casebase}. Next, we provide some details
about the implementation of case-base sampling in our package, and we
give a brief survey of its main functions. This is followed by three
case studies that illustrate the flexibility and capabilities of
\pkg{casebase}. We show how the same framework can be used for
non-linear functions of time, competing risks and variable selection via
penalized regression. Finally, we end the article with a discussion of
the results and of future directions.

@@

\hypertarget{comparison-with-existing-packages}{%
\section{Comparison with existing
packages}\label{comparison-with-existing-packages}}

@@

Parametric survival models are implemented in a handful of packages,
each differing in the underlying data distribution: \pkg{CFC}
\citeyearpar{mahani2015bayesian}, \pkg{flexsurv} \citeyearpar{flexsurv},
\pkg{SmoothHazard} \citeyearpar{smoothHazard}, \pkg{rsptm2}
\citeyearpar{clements_liu}, \pkg{mets}
\citeyearpar{scheike2014estimating}, and \pkg{survival}
\citeyearpar{survival-package}. For example, \pkg{SmoothHazard} is
limited to Weibull distributions \citeyearpar{smoothHazard}, whereas
both \pkg{flexsurv} and \pkg{survival} allow users to supply any
distribution of their choice. \pkg{flexsurv}, \pkg{smoothhazard},
\pkg{mets} and \pkg{rstpm2} can model the effect of time using splines,
which allows flexible modeling of the hazard function. As discussed
above, \pkg{casebase} can model any parametric family whose log-hazard
can be expressed as a linear combination of covariates (including time).
Therefore, our package allows the user to model the effect of time using
splines. Also, by including interaction terms between covariates and
time, it also allows users to fit (non-proportional) time-varying
coefficient models. However, unlike \pkg{flexsurv}, we do not explicitly
model any shape parameter.
